A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/Examples.lagda.tex
revised examples

In this section we illustrate a few simple examples of processes encoded in our
formalization. We start by encoding the type of a boolean value being produced
as a choice between ``going left'' (which we interpret as true) and ``going
right'' (which we interpret as false).

@@

The boolean constants are processes that perform one of these choices.

\begin{code}
true : Proc [ 𝔹 ]
true = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))

false : Proc [ 𝔹 ]
false = select (ch ⟨ < ≫ ⟩ inj₂ (close ch))
\end{code}

We can use the host language for programming higher-order processes. For
example, we can define a conditional process thus:

\begin{code}
if_else : ∀{Γ} → Proc Γ → Proc Γ → Proc (dual 𝔹 ∷ Γ)
if P else Q = curry∗ case ch (< ≫) ( wait (ch ⟨ < ≫ ⟩ P)
                                   , wait (ch ⟨ < ≫ ⟩ Q))
\end{code}

A term $\AgdaFunction{if}~P~\AgdaFunction{else}~Q$ is a process that waits for a
boolean value (\cf the $\AgdaFunction{dual}~\mathbb{B}$ type at the front of its
typing context) and continues as either $P$ or $Q$ depending on whether it
receives true or false. We use \AgdaFunction{curry∗} (defined in
\Cref{sec:context-agda}) to curry the constructor
\AgdaInductiveConstructor{case} so that we can supply its arguments one by one
saving a few parentheses and reducing clutter.

Next we define a process $\AgdaFunction{drop}~P$ that consumes a boolean and
continues as $P$ regardless of its value.

\begin{code}
drop : ∀{Γ} → Proc Γ → Proc (dual 𝔹 ∷ Γ)
drop P = if P else P
\end{code}

@@

\begin{code}
!!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
!!_ B = curry∗ cut B ≫ (if false else true)

_&&_ _||_  : Proc [ 𝔹 ] → Proc [ 𝔹 ] → Proc [ 𝔹 ]
A && B   =  curry∗ cut A ≫ $
            curry∗ cut B ≫ $
            if (curry∗ link ch (< ≫) ch) else (drop false)
A || B   = !! ((!! A) && (!! B))
\end{code}

@@

\begin{code}
{-# TERMINATING #-}
eval : ∀{Γ} → Proc Γ → Proc Γ
eval P with deadlock-freedom P
... | inj₁ (Q , _ , _)  = Q
... | inj₂ (Q , _)      = eval Q
\end{code}

Now if we ask Agda to normalize the goal
$\AgdaFunction{eval}~(\AgdaFunction{false}~\AgdaFunction{||}~\AgdaFunction{false})$
we obtain
$\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
that is the definition of \AgdaFunction{false}, as expected.

For the example that follows, it is convenient to introduce the linear
implication connective according to the standard definition $A \multimap B
\eqdef \dual{A} \Par B$.

@@

Below we define a persistent, polymorphic echo \AgdaInductiveConstructor{server}
that waits for a value and sends it back unchanged. The server is polymorphic in
the sense that the type of the value is arbitrary and denoted by the type
variable $X$. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
library and creates an element of \AgdaDatatype{Fin} from the natural number
$n$.

\begin{code}
echo : let X = var (# 0) in
       Proc [ `! (`∀ (X ⊸ X)) ]
echo = curry∗ server ch (< ≫)
             ( un-[]
             , curry∗ all ch (< ≫) λ X →
               curry∗ join ch (< ≫) $
               curry∗ link ch (< ≫) ch)
\end{code}

@@

\begin{code}
echo-true : Proc [ 𝔹 ]
echo-true = curry∗ cut echo ≫ $
            curry∗ client ch (< ≫) $
            curry∗ ex ch (< ≫) $
            curry∗ fork ch (< ≫) $ true ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
\end{code}

If we ask Agda to evaluate this process we obtain \AgdaFunction{true}, as
expected.

From these examples it should be clear that writing \Calculus processes in Agda
is not entirely straightforward, but it is still doable with a little effort.
The lack of names (for channels) clearly obfuscates the code, but the context
splittings that must be provided are usually small and fairly easy to
understand. Most importantly, the use of \AgdaFunction{≫} helps truncating
context splittings as soon as the seeked channel has been reached.
%
Either way, the formalization is meant to study the meta-theory of \Calculus
rather than implementing a user-friendly DSL for it.

>>>>

In this section we revisit and expand the sample processes discussed in
\Cref{ex:booleans,ex:echo} and show their encoding in our formalization. The
encoding of $\Bool$ is straightforward.

@@

The boolean constants are encoded thus:
\begin{code}
True : Proc [ 𝔹 ]
True = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))

False : Proc [ 𝔹 ]
False = select (ch ⟨ < ≫ ⟩ inj₂ (close ch))
\end{code}

We can take advantage of the host language for programming higher-order
processes. For example, we can define a conditional process thus:

\begin{code}
If_Else : ∀[ Proc ⇒ Proc ⇒ (dual 𝔹 ∷_) ⊢ Proc ]
If P Else Q = curry∗ case ch (< ≫) ( wait (ch ⟨ < ≫ ⟩ P)
                                   , wait (ch ⟨ < ≫ ⟩ Q))
\end{code}

A term $\AgdaFunction{If}~P~\AgdaFunction{Else}~Q$ is a process that waits for a
boolean value (\cf the $\AgdaFunction{dual}~\mathbb{B}$ type at the front of its
typing context) and continues as either $P$ or $Q$ depending on whether it
receives true or false. We use \AgdaFunction{curry∗} (defined in
\Cref{sec:context-agda}) to curry the constructor
\AgdaInductiveConstructor{case} so that we can supply its arguments one by one
saving a few parentheses and reducing clutter.

Next we define a process $\AgdaFunction{Drop}~P$ that consumes a boolean and
continues as $P$ regardless of its value.

\begin{code}
Drop : ∀[ Proc ⇒ (dual 𝔹 ∷_) ⊢ Proc ]
Drop P = If P Else P
\end{code}

@@

\begin{code}
!!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
!!_ B = curry∗ cut B ≫ (If False Else True)

_&&_ _||_  : Proc [ 𝔹 ] → Proc [ 𝔹 ] → Proc [ 𝔹 ]
A && B   =  curry∗ cut A ≫ $
            curry∗ cut B ≫ $
            If (curry∗ link ch (< ≫) ch) Else (Drop False)
A || B   = !! ((!! A) && (!! B))
\end{code}

@@

\begin{code}
{-# TERMINATING #-}
eval : ∀[ Proc ⇒ Proc ]
eval P with deadlock-freedom P
... | inj₁ (Q , _ , _)  = Q
... | inj₂ (Q , _)      = eval Q
\end{code}

Now if we ask Agda to normalize the goal
$\AgdaFunction{eval}~(\AgdaFunction{False}~\AgdaFunction{||}~\AgdaFunction{False})$
we obtain
$\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
that is the definition of \AgdaFunction{False}, as expected.

For the encoding of the echo server (\Cref{ex:echo}), it is convenient to
introduce the linear implication connective according to the standard definition
$A \multimap B \eqdef \dual{A} \Par B$.

@@

Recall that the server is polymorphic in the sense that the type of the value it
receives and sends back is arbitrary and denoted by a type variable $X$. The
notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and creates
an element of \AgdaDatatype{Fin} from the natural number $n$.

\begin{code}
Echo : let X = var (# 0) in
       Proc [ `! (`∀ (X ⊸ X)) ]
Echo = curry∗ server ch (< ≫)
             ( un-[]
             , curry∗ all ch (< ≫) λ X →
               curry∗ join ch (< ≫) $
               curry∗ link ch (< ≫) ch)
\end{code}

@@

\begin{code}
Echo-True : Proc [ 𝔹 ]
Echo-True = curry∗ cut Echo ≫ $
            curry∗ client ch (< ≫) $
            curry∗ ex ch (< ≫) $
            curry∗ fork ch (< ≫) $ True ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
\end{code}

If we ask Agda to evaluate this process we obtain \AgdaFunction{True}, as
expected.

From these examples it should be clear that writing \Calculus processes in Agda
is not entirely straightforward and requires a little effort. The lack of names
(for channels) clearly obfuscates the code, but the context splittings that must
be provided are usually small and fairly easy to understand. Most importantly,
the use of \AgdaFunction{≫} helps truncating context splittings as soon as the
seeked channel has been reached.
%
Either way, the formalization is meant to study the meta-theory of \Calculus
rather than implementing a user-friendly DSL for it.
